
\documentclass{article}
\include{preamble.sty}

\begin{document}

  \section*{\sc Assignment \#7: YouTube Short}

    Your task is to create (independently or with a partner) a 15-to-60-second video related to sport analytics.
  
    \subsection*{\sc Why are you being asked to do this?}

      In today's world, there are other (perhaps better?) ways to reach people with your ideas than publishing a research paper. Let's flex our creativity and experiment with one alternative medium. Plus, if you allow the department to post your video, you will be contributing to the department's recruiting efforts! Maybe one day you'll meet a younger student who was inspired to study sport analytics because of your video.

    \subsection*{\sc What (exactly) are you being asked to do?}

      Create a short video about some topic in sport analytics. If it's about a topic we've covered in class, great! But that's not a requirement. Maybe it's educational. Maybe it's funny. Maybe it's both!
      
      \subsubsection*{\sc Submission Requirements}

        \begin{itemize}
          \item A 15-to-60-second video formatted as a MOV file
          \item If you do not want your video to be published online, please indicate that in the text entry field of the assignment submission page. Otherwise (pending department approval), your video will be published as a short on the Rice Sport Management YouTube channel.
        \end{itemize}

    \subsection*{\sc How will your grade be determined?}
      
    This assignment will be graded for completion, meaning that as long as you complete the assignment, your grade will be 100\%.

\end{document}